Global patterns and drivers of vertebrate verticality

### 2024-08-16 verticality_future_mkd.Rmd
full first draft

---
date: "`r Sys.Date()`"
author: "Your Name"
title: "officedown template"
bibliography: "`r rbbt::bbt_write_bib('docs/r-refs.bib', overwrite = TRUE)`"
output: 
  officedown::rdocx_document:
    mapstyles:
      Normal: ['First Paragraph']
plots:
  caption:
    style: Image Caption
    pre: 'Figure '
    sep: ': '
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.cap = TRUE, include=FALSE, warning=FALSE, message=FALSE, error=FALSE)
library(officedown)
library(officer)
library(tidyverse)
library(spatialreg)
library(data.table)
library(flextable)
library(knitr)

@@

ft <- fp_text(shading.color='#EFEFEF', bold = TRUE)
```

@@

Vertical stacking of species is an emblematic signature of biologically rich regions and is associated with patterns of species richness from local to global scales [@macarthurPopulationEcologyWarblers1958; @macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019; @bashamVerticalStratificationPatterns2023a]. At local scales, species assemblages may be stratified across vertical resource and climate gradients from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023]. Tall, multi-layered forests with high foliage height diversity increase both the volume of habitat and the niche opportunities within the habitat, acting effectively as multi-floored apartments that enable species to vertically partition resources [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. By reducing competition, vertical stratification may thus act as a mechanistic link to positive relationships between canopy height and species richness [@wagnerRemotelySensedTree2023] (*cite positive relationships between canopy height and species richness at local scales*). *concept fig* This hypothesis has been supported by positive relationships between verticality, defined here as the average vertical position of species in an assemblage, and species richness [@oliveiraVerticalStratificationInfluences2019]. However, canopy height and other metrics of vegetation structure do not consistently impact species richness at regional to global scales, with notable declines in the strength of the relationship at higher latitudes where the diversity of arboreal species generally declines [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019]. Examining the impacts of climatic and vegetative drivers of vertical stratification may thus add nuance to our understanding of the impact of canopy height on species richness.

@@

Spatial variation may also occur in the relationship between species richness and  verticality [@oliveiraVerticalStratificationInfluences2019]. Though prior studies have shown positive relationships between verticality and richness, the strength of this relationship is variable and generally more positive in tropical regions [@oliveiraVerticalStratificationInfluences2019] *additional citations for assumed relationship*. Where climatic conditions allow canopy height to positively impact verticality, as may be expected in the wet tropics, the positive relationship between verticality and richness may be stronger because the addition of a vegetation layer allows the addition of arboreal species to the assemblage. In contrast, where taller canopies do not facilitate higher verticality, the relationship between verticality and richness may be weak (*concept fig*).

@@

$$
y_s ∼ 𝜇_{s}, \\
𝜇_s = 𝑓^{-1}(𝜲^{main}_s𝛽 + α_g + 𝜲^{svc}_sζ_s + ⍵_s), \\
⍵_s ∼ MVNormal(0,𝚺_⍵)
$$

@@

*Calculate relative change as (future-present)/(presentmax-presentmin)*

```{r}
# calculate percent change in verticality
#source("./../scripts/04_plots/vertdif_by_biorealm.R")
source("scripts/04_plots/vertdif_by_biorealm.R")

@@

Climate influences global patterns of verticality because more extreme climates in thh canopy, including increased exposure to hot, cold, and dry conditions, pose physiological challenges to living in arboreal habitats [@oliveiraVerticalStratificationInfluences2019]. Water availability therefore facilitates verticality, while temperatures at both hot and cold extremes and water scarcity generally limit verticality. For example, more severe water limitation in the canopy than the understory [@scheffersIncreasingArborealityAltitude2013] may contribute to observed declines in arboreal life strategies outside moist forests in tropical regions. However, our models indicated nuances in the relative influences of different drivers across taxa.

@@

Increasing temperature and declining precipitation could alter global patterns of assemblage verticality via vertical niche shifts, range shifts, or local extinction. Our models indicated relatively small declines of verticality in tropical latitudes in response to climate change at a global scale. In the absence of climatic buffering experienced in the understory, species living in arboreal habitats may be threatened by the increasing magnitude and frequency of extreme heat and drought [@ipccClimateChange20232023]. The largest predicted declines in verticality occurred for amphibians in the tropics, corresponding with small themral safety margins of tropical relative to temperate ectotherms [@sundayThermalsafetyMarginsNecessity2014]. To avoid unsuitable conditions that exceed physiological tolerances, species could behaviorally regulate climate exposure by spending more time in the understory, thus reducing assemblage verticality [@sundayThermalsafetyMarginsNecessity2014; @bashamVerticalStratificationCollapses2020]. Indeed, vertical niche shifts have been observed at smaller spatiotemporal scales across tropical forests. For example, birds and arboreal frogs increase their use of understory resources during the dry season [@delgado-martinezDifferentialUtilizationSurface2023; @bashamVerticalStratificationCollapses2020]. Arboreal primates from the Americas to Madagascar similarly increase their use of terrestrial habitats under hotter temperatures [@eppleyFactorsInfluencingTerrestriality2022]. Alternatively, arboreal species could become locally extinct if they are not able to adjust microhabitat use to avoid extreme climate conditions.

@@

```{r fig.id="vert-map-global", fig.width=160, fig.height=96, fig.cap="SES and mean verticality of birds, mammals, reptiles, and amphibians. Line plots inidicate the average proportion of arboreal (red), terrestrial (yellow), and fossorial (blue) species across latitudes. Gray areas indicate areas that are not wooded or that have fewer than five species."}

read_docx() |>
  body_add_img(src = "figures/main_figs/vert_maps.png", style="centered", width=160, height=96)
#include_graphics("figures/main_figs/vert_maps.png")
```

```{r fig.id="mod-coefs", fig.width=160, fig.height=87.3, fig.cap="Coefficient estimates for models of mean and SES verticality with 95% confidence intervals. tmin_cold = minimum temperature of the coldest month, tmax_warm = maximum temperature of the warmest month, precip_dry = precipitation of the driest month, precip_wet = precipitation of the wettest month, veg_den = vegetation density, clim_velocity = climate velocity"}

read_docx() |>
  body_add_img(src = "figures/main_figs/sdmTMB_coefs.png", style="centered", width=160, height=87.3)

#include_graphics("figures/main_figs/sdmTMB_coefs.png", error = F)
```

```{r fig.id="svc", fig.width=160, fig.height=143.95, fig.cap="Spatially varying coefficients for canopy height in models GLMMs of SES and mean verticality for birds, mammals, reptiles, and amphibians. Gray areas indicate areas that are not wooded or that have fewer than five species."}

#include_graphics("figures/main_figs/canopy_height_coefs.png", error = F)
```

```{r fig.id="corplot", fig.width = 160, fig.height=120, fig.cap="Correlation between richness and SES verticality (A-D) or mean verticality (E-H) for all points and for points where the canopy height coefficient was significant and positive. "}

#include_graphics("figures/main_figs/vert_rich_cor.png", error = F)
```

```{r fig.id="vert-dif-map", fig.width=160, fig.height=133.3, fig.cap="Relative percent change in SES and mean verticality for birds, mammals, reptiles, and amphibians. Bar plots inidcate average relative percent change per latitude. Gray areas indicate areas that are not wooded or that have fewer than five species."}

#include_graphics("figures/main_figs/vert_difs.png", error = F)
```

@@

@@

```{r fig.id="vert-biome", fig.lp="ed", fig.cap="", fig.height=80, fig.width=160}
#include_graphics("figures/supp_figs/vert_biome_forestOnly_forestSES.png", error = F)
```

```{r fig.id="vert-dif-biome", fig.lp="ed", fig.cap="Percent relative difference between SES and mean verticality predicted under present and future climate conditions for global biomes. Negative values indicate predicted declines in verticality and positive values indicate predicted increases in verticality. Biomes along the y-axis are ordered from those that were predicted to exhibit the greatest average in verticality across taxa at the top to those that were predicted to exhibit the greatest average decrease in verticality at the bottom.", fig.height=80, fig.width=160}

#include_graphics("figures/supp_figs/vert_biome_forestOnly_forestSES.png", error = F)
```

```{r fig.id="svc-sig", fig.lp="ed", fig.cap="Areas in which the spatially varying coefficient for canopy height had a significant positive or negative effect on SES or mean verticality of birds, mammals, reptiles, and amphibians.", fig.height=143.9, fig.width=160}

#include_graphics("figures/supp_figs/svc_canopyHeight_sig_link.png", error = F)
```

>>>>

---
author: "Lydia Soifer, David Edwards, Brett Scheffers"
title: "Global patterns and drivers of vertebrate verticality"
bibliography: "`r rbbt::bbt_write_bib('docs/r-refs.bib', overwrite = TRUE)`"
output: 
  officedown::rdocx_document:
    reference_docx: "style-guide.docx"
    mapstyles:
      Normal: ['First Paragraph']
plots:
  caption:
    style: Image Caption
    pre: 'Figure '
    sep: ': '
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.cap = TRUE, warning=FALSE, message=FALSE, error=FALSE)
library(officedown)
library(officer)
library(tidyverse)
library(spatialreg)
library(data.table)
library(flextable)
library(knitr)
library(chunkhooks)

@@

ft <- fp_text(shading.color='#EFEFEF', bold = TRUE)

hook_figure_unit("mm")
```

@@

Vertical stacking of species is an emblematic signature of biologically rich regions and is associated with patterns of species richness from local to global scales [@macarthurPopulationEcologyWarblers1958; @macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019; @bashamVerticalStratificationPatterns2023a]. At local scales, species assemblages may be stratified across vertical resource and climate gradients from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023]. Tall, multi-layered forests with high foliage height diversity increase both the volume of habitat and the niche opportunities within the habitat, acting effectively as multi-floored apartments that enable species to vertically partition resources [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. By reducing competition, vertical stratification may thus act as a mechanistic link to positive relationships between canopy height and species richness [@culbertInfluenceVerticalHorizontal2013; @fengForestCanopyHeight2020; @gillespieDistributionBirdSpecies2001; @wagnerRemotelySensedTree2023]. *concept fig* This hypothesis has been supported by positive relationships between verticality, defined here as the average vertical position of species in an assemblage, and species richness [@oliveiraVerticalStratificationInfluences2019]. However, canopy height and other metrics of vegetation structure do not consistently impact species richness at regional to global scales, with notable declines in the strength of the relationship at higher latitudes where the diversity of arboreal species generally declines [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019]. Examining the impacts of climatic and vegetative drivers of vertical stratification may thus add nuance to our understanding of the impact of canopy height on species richness.

@@

Spatial variation may also occur in the relationship between species richness and  verticality [@oliveiraVerticalStratificationInfluences2019]. Though prior studies have shown positive relationships between verticality and richness, the strength of this relationship is variable and generally more positive in tropical regions [@cabralFunctionalTraitsPhylogeny2022; @oliveiraVerticalStratificationInfluences2019]. Where climatic conditions allow canopy height to positively impact verticality, as may be expected in the wet tropics, the positive relationship between verticality and richness may be stronger because the addition of a vegetation layer allows the addition of arboreal species to the assemblage. In contrast, where taller canopies do not facilitate higher verticality, the relationship between verticality and richness may be weak (*concept fig*).

@@

$$
\begin{align}
y_s ∼ 𝜇_{s}, \\
𝜇_s = 𝑓^{-1}(𝜲^{main}_s𝛽 + α_g + 𝜲^{svc}_sζ_s + ⍵_s), \\
⍵_s ∼ MVNormal(0,𝚺_⍵)
\end{align}
$$

@@

```{r cache=TRUE, cache.path="cache/"}
# calculate percent change in verticality
#source("./../scripts/04_plots/vertdif_by_biorealm.R")
source("scripts/04_plots/vertdif_by_biorealm.R")

@@

Climate influences global patterns of verticality because more extreme climates in the canopy, including increased exposure to hot, cold, and dry conditions, pose physiological challenges to living in arboreal habitats [@oliveiraVerticalStratificationInfluences2019]. Water availability therefore facilitates verticality, while temperatures at both hot and cold extremes and water scarcity generally limit verticality. For example, more severe water limitation in the canopy than the understory [@scheffersIncreasingArborealityAltitude2013] may contribute to observed declines in arboreal life strategies outside moist forests in tropical regions. However, our models indicated nuances in the relative influences of different drivers across taxa.

@@

Increasing temperature and declining precipitation could alter global patterns of assemblage verticality via vertical niche shifts, range shifts, or local extinction. Our models indicated relatively small declines of verticality in tropical latitudes in response to climate change at a global scale. In the absence of climatic buffering experienced in the understory, species living in arboreal habitats may be threatened by the increasing magnitude and frequency of extreme heat and drought [@ipccClimateChange20232023]. The largest predicted declines in verticality occurred for amphibians in the tropics, corresponding with small thermal safety margins of tropical relative to temperate ectotherms [@sundayThermalsafetyMarginsNecessity2014]. To avoid unsuitable conditions that exceed physiological tolerances, species could behaviorally regulate climate exposure by spending more time in the understory, thus reducing assemblage verticality [@sundayThermalsafetyMarginsNecessity2014; @bashamVerticalStratificationCollapses2020]. Indeed, vertical niche shifts have been observed at smaller spatiotemporal scales across tropical forests. For example, birds and arboreal frogs increase their use of understory resources during the dry season [@delgado-martinezDifferentialUtilizationSurface2023; @bashamVerticalStratificationCollapses2020]. Arboreal primates from the Americas to Madagascar similarly increase their use of terrestrial habitats under hotter temperatures [@eppleyFactorsInfluencingTerrestriality2022]. Alternatively, arboreal species could become locally extinct if they are not able to adjust microhabitat use to avoid extreme climate conditions.

@@

```{r fig.id="vert-map-global", fig.width=160, fig.height=96, fig.cap.pre="Fig. ", fig.cap="SES and mean verticality of birds, mammals, reptiles, and amphibians. Line plots inidicate the average proportion of arboreal (red), terrestrial (yellow), and fossorial (blue) species across latitudes. Gray areas indicate areas that are not wooded or that have fewer than five species.", include=TRUE}

knitr::include_graphics("figures/main_figs/vert_maps.png", error = F)
```

```{r fig.id="mod-coefs", fig.width=160, fig.height=87.3, fig.cap.pre="Fig. ", fig.cap="Coefficient estimates for models of mean and SES verticality with 95% confidence intervals. tmin_cold = minimum temperature of the coldest month, tmax_warm = maximum temperature of the warmest month, precip_dry = precipitation of the driest month, precip_wet = precipitation of the wettest month, veg_den = vegetation density, clim_velocity = climate velocity"}

include_graphics("figures/main_figs/sdmTMB_coefs.png", error = F)
```

```{r fig.id="svc", fig.width=160, fig.height=143.95, fig.cap.pre="Fig. ", fig.cap="Spatially varying coefficients for canopy height in models GLMMs of SES and mean verticality for birds, mammals, reptiles, and amphibians. Gray areas indicate areas that are not wooded or that have fewer than five species."}

include_graphics("figures/main_figs/canopy_height_coefs.png", error = F)
```

```{r fig.id="corplot", fig.width = 160, fig.height=213.4, fig.cap.pre="Fig. ", fig.cap="Correlation between richness and SES verticality (A-D) or mean verticality (E-H) for all points and for points where the canopy height coefficient was significant and positive. "}

include_graphics("figures/main_figs/vert_rich_cor.png", error = F)
```

```{r fig.id="vert-dif-map", fig.width=160, fig.height=133.3, fig.cap.pre="Fig. ", fig.cap="Relative percent change in SES and mean verticality for birds, mammals, reptiles, and amphibians. Bar plots inidcate average relative percent change per latitude. Gray areas indicate areas that are not wooded or that have fewer than five species."}

include_graphics("figures/main_figs/vert_reldifs.png", error = F)
```

@@

# Supplementary methods
## Species data

We identified wooded habitats by ecoregions including forests, woodlands, taiga, várzea, Yungas, savanna, thicket, and mallee as defined by Dinnerstein et al. (2017).

The TetrapodTraits dataset uses phylogenetic imputation to identify traits if trait data is not available in the literature. For species where imputation was necessary, we considered a species to use a given vertical stratum if the imputation probability exceeded 0.7.

## Mesh development

@@

```{r fig.id="vert-biome", fig.lp="suppfig", fig.cap="", fig.cap.pre="Fig. S", fig.autonum.start_at=1, fig.height=80, fig.width=160}
include_graphics("figures/supp_figs/vert_biome_forestOnly_forestSES.png", error = F)
```

```{r fig.id="vert-dif-biome", fig.lp="suppfig", fig.cap.pre="Fig. S", fig.cap="Percent relative difference between SES and mean verticality predicted under present and future climate conditions for global biomes. Negative values indicate predicted declines in verticality and positive values indicate predicted increases in verticality. Biomes along the y-axis are ordered from those that were predicted to exhibit the greatest average in verticality across taxa at the top to those that were predicted to exhibit the greatest average decrease in verticality at the bottom.", fig.height=80, fig.width=160}

include_graphics("figures/supp_figs/vert_biome_forestOnly_forestSES.png", error = F)
```

```{r fig.id="svc-sig", fig.lp="suppfig", fig.cap.pre="Fig. S", fig.cap="Areas in which the spatially varying coefficient for canopy height had a significant positive or negative effect on SES or mean verticality of birds, mammals, reptiles, and amphibians.", fig.height=143.9, fig.width=160}

include_graphics("figures/supp_figs/svc_canopyHeight_sig_link.png", error = F)
```
